---
title: "Reproducible Research: Peer Assessment 1"
author: "Georges Bressange"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
The zip file, as stated in the peer assesment instructions, is already 
present in the working directory. We unzip it:

```{r}
unzip("activity.zip")
```

Then, we read the zip file to produce the analytic data frame we will 
later use to perform statistic analysis:

```{r}
activity <- read.csv("activity.csv")
```

We convert the factor "date" to a date format:

```{r}
activity$date <- as.Date(activity$date, format = "%Y-%m-%d")
```
Let us visualize now the structure of the "activity" data frame:

```{r}
str(activity)
```
## What is mean total number of steps taken per day?
Let us create a new variable "days" in the data frame "activity" to separate the 
observations along days:

```{r}
activity$Days <- as.Date(cut(activity$date, breaks = "days"))
```

Let us visualize the structure of "activity":

```{r}
summary(activity)
```
We now plot an histogram of the total number of steps taken each day using 
ggplot2 (we have interpreted "each day" as "each of the 61 days" since the 
instructions don't ask for "each day of the week"). NA values have been 
ignored.

```{r}
library(ggplot2)
g <- ggplot(data = activity, aes(Days, steps))
g + stat_summary(fun.y = sum, geom = "bar", na.rm = TRUE)
```
 
We now report the mean and median total number of steps taken per day. We first 
load the dplyr package:
 
```{r}
library(dplyr)
```

and summarize with the following naming: the mean total number of steps per day is 
"mean.steps.daily" and the  the median total number of steps per day is 
"median.steps.daily":

```{r}
df1 <- summarise(activity, mean.steps.daily = mean(steps, na.rm = TRUE),
                 median.steps.daily = median(steps, na.rm = TRUE))
df1
```

## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
