---
title: "Reproducible Research: Peer Assessment 1"
author: "Georges Bressange"
output: 
  html_document:
    keep_md: yes
    toc: yes
---


## Loading and preprocessing the data
1. The zip file, as stated in the peer assesment instructions, is already 
present in the working directory. We unzip it:

```{r}
unzip("activity.zip")
```

Then, we read the zip file to produce the analytic data frame we will 
later use to perform statistic analysis:

```{r}
activity <- read.csv("activity.csv")
```

```{r}
str(activity)
```

2. We convert the factor "date" to a date format:

```{r}
activity$date <- as.Date(activity$date, format = "%Y-%m-%d")
```

```{r}
str(activity)
```

The "interval" variable is not an arithmetic suite of time: after the 55 minute interval, the id interval jumps to 100 as we can see from the command:

```{r}
unique(activity$interval)
```

The series of time intervals and those we would like to have are then:

| Series of 5 min intervals   | New Series of 5 min intervals|
| ----------------------------| ----------------------------|
| 0 - 5 - 10 -  ...      - 55 | 0 - 5 - 10 -  ...      - 55 |
| 100 - 105 - 110 - ... - 155 | 60 - 65 - 70 - ...    - 115 |
| 200 - 205 - 210 - ... - 255 | 120 - 125 - 130 - ... - 175 |
| 300 - 305 - 310 - ... - 355 | 180 - 185 - 190 - ... - 235 |
| ........................... | ........................... |
| 2300 - 2305 - ... -    2355 | 1380 - 1385 - ... -    1435 |

A mathematical function to make this transformation is

```{r}
fint <- function(x){x - 40*floor(x/100)}
```

where floor(y) is the integer part of a real number y.

Using the mutate function of the dplyr library, let us transform the interval variable according to previous function:

```{r}
library(dplyr)
```
```{r}
activity <- mutate(activity, interval = fint(interval))
```

```{r}
unique(activity$interval)
```

We see now that the last 5 minute interval is 1435-1440 which is coherent since in a day there are 1440 minutes (24 times 60).

## What is mean total number of steps taken per day?
1. Let us calculate the total number of steps taken per day in "activity":

```{r}
library(dplyr)
```

```{r}
totalnumbersteps1 <- summarise(group_by(activity, date), total.steps.daily = sum(steps, na.rm = TRUE))
```

Let us present the head of the table of results:

```{r}
head(totalnumbersteps1)
```

2. We now plot an histogram of the total number of steps taken each day using 
ggplot2 (we have interpreted "each day" as "each of the 61 days" since the 
instructions don't ask for "each day of the week"). NA values have been 
ignored. A word on the differences between histogram and barplot: an histogram 
is used to describe the ditribution of one quantitative variable. A barplot is used 
to compare several categorical variables. Here, we want to study the distribution 
of the "steps" variable but this variable is partially already counted so we 
have used a barplot to be able to specify the counted steps on the y-axis.

```{r}
library(ggplot2)
```

```{r, fig.show='asis'}
g <- ggplot(data = activity, aes(date, steps))
g + stat_summary(fun.y = sum, geom = "bar", na.rm = TRUE) +
        ggtitle("Total number of steps taken each day")
```
 
3. We now report the mean and median total number of steps taken per day. The dplyr 
package has already been loaded. We summarize with the following naming: 
the mean total number of steps per day is "mean.steps.daily" and the  the median 
total number of steps per day is "median.steps.daily":

```{r}
report1 <- summarise(activity, mean.steps.daily = mean(steps, na.rm = TRUE),
                 median.steps.daily = median(steps, na.rm = TRUE))
report1
```

The anonymous individual performs around `r round(report1$mean.steps.daily[1])` steps 
per day and the fact that the median total number of steps per day is zero means 
that for at least 50% of the days in these two months, the anonymous individual 
don't perform any step (or no steps were recorded 50% of the time).

## What is the average daily activity pattern?
1. We group similar interval of 5 mins along the days. The resulting data frame 
is called "activity_by_intervals".

```{r}
activity_by_intervals <- aggregate(activity, by = list(activity$interval), 
                             FUN = "mean", na.rm = TRUE)
activity_by_intervals <- select(activity_by_intervals, interval, steps)
activity_by_intervals <- rename(activity_by_intervals, 
                                average.steps.by.interval = steps)
```

```{r}
summary(activity_by_intervals)
```

We now make a time series plot (type = "l") of the 5-minute interval (x-axis) and 
the average number of steps taken, averaged across all days (y-axis)

```{r}
plot(activity_by_intervals$interval, activity_by_intervals$average.steps.by.interval, 
     type = "l", xlab = "intervals of 5 min", 
     ylab = "Average total steps per day", 
     main ="Time series average total steps \n per day by interval")
```

2. The 5-minute interval, on average across all the days in the dataset, 
contains the maximum number of steps can be obtained by the following 
commands, performed on the dataset "activity_by_intervals":

```{r}
M <- max(activity_by_intervals$average.steps.by.interval) ## calculates the maximum average steps in the series of intervals
n <- activity_by_intervals[
        activity_by_intervals$average.steps.by.interval == M,]$interval ## gives the time interval where this maximum occurs
activity_by_intervals[activity_by_intervals$average.steps.by.interval == M,]
```

The 5-minute interval which, on average across all the days in the dataset, 
that contains the maximum number of steps is the interval `r n` i.e., after 
time conversion, between 08h00 and 08h05. To confirm that, we ploted again 
the time series presented in 1. by adding a blue horizontal line 
corresponding to the maximum of steps (y = `r M`) end a red vertical line 
corresponding to the interval where this maximum occurs (x = `r n`).

```{r}
plot(activity_by_intervals$interval, activity_by_intervals$average.steps.by.interval, 
     type = "l", xlab = "intervals of 5 min", 
     ylab = "Average total steps per day", 
     main ="Time series of average total steps \n per day by interval")
abline(h = M, col = "blue")
abline(v = n, col = "red")
text(x = 650, y = 195, labels = "y = 206.17", col = "blue")
text(x = 650, y = 125, labels = "x = 515", col = "red")
```

## Imputing missing values
1. Calculate and report the total number of missing values in the dataset 

```{r}
nasteps <- sum(is.na(activity$steps)) # number of NA in steps column
nadate <- sum(is.na(activity$date)) # number of NA in date column
nainterval <- sum(is.na(activity$interval)) # number of NA in interval column
nas <- data.frame(nasteps, nadate, nainterval)
nas
```

The dataset "activity" contains `r nasteps` missing values in total. These 
NA values are exclusively located in the "steps" column. This number of missing values can also be read from the summary command applied to "activity":

```{r}
summary(activity)
```

2. Strategy for filling in all of the missing values in the dataset. 

We decide to replace each missing "steps"" value by the mean of the 5-minute interval 
to which they belong.

3. Creating a new dataset "activityfill" that is equal to the original dataset but with the missing data filled in.

```{r}
activityfill <- activity ## creating a new dataset
listNAintervals <- activityfill$interval[is.na(activityfill$steps)] ## list the values of time intervals corresponding to the missing steps values
f <- function(x){
        subset(activity_by_intervals, interval == x)$average.steps.by.interval
} 
activityfill$steps[is.na(activityfill$steps)] <- sapply(listNAintervals, f) ## the function f will replace each missing steps value in "activity" by the corresponding average.steps.by.interval extracted from "activity_by_intervals".
```

4. Histogram of the total number of steps taken each day:
We use the same plot command as used for "activity" dataset.

```{r}
g <- ggplot(data = activityfill, aes(date, steps))
g + stat_summary(fun.y = sum, geom = "bar", na.rm = TRUE) + 
        ggtitle("Total number of steps taken each day")
```

We now report the mean and median total number of steps taken per day. 

```{r}
report2 <- summarise(activityfill, mean.steps.daily = mean(steps, na.rm = TRUE),
                 median.steps.daily = median(steps, na.rm = TRUE))
report2
```

These values don't differ from the estimates from the first part of the assignment.

Let us calculate the total number of steps taken per day in "activityfill":

```{r}
totalnumbersteps2 <- summarise(group_by(activityfill, date), total.steps.daily = sum(steps, na.rm = TRUE))
head(totalnumbersteps2)
```

As a result of imputing missing values, we calculate below the average of the increase of the total number of steps per day:

```{r}
impactfill <- data.frame(date = totalnumbersteps1$date, 
                         daily.increase = totalnumbersteps2$total.steps.daily -    
                                 totalnumbersteps1$total.steps.daily)
report3 <- summarise(impactfill, mean.daily.increase = mean(daily.increase), 
          median.daily.increase = median(daily.increase))
meanincreasesteps <- report3$mean.daily.increase
head(report3)
```

Therefore, we see that in average, the daily total number of steps has increased from 
around `r round(meanincreasesteps)` steps.

## Are there differences in activity patterns between weekdays and weekends?

1. We create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

The following code is useful because my locales are french and I want to obtain weekdays in english.

```{r}
curr_locale <- Sys.getlocale("LC_TIME")
Sys.setlocale("LC_TIME","en_US.UTF-8")
```

Creating the weekdays factor variable:

```{r}
weekdayslist <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')
activityfill$weekDay <- factor((weekdays(activityfill$date) %in% weekdayslist), 
                               levels=c(FALSE, TRUE), labels=c('weekend', 'weekday')) 
```

Let us group similar interval of 5 mins along the days. The resulting data frame 
is called "activityfill_by_intervals".

```{r}
by_weekdays_and_interval <- group_by(activityfill, weekDay, interval)
activityfill_by_intervals <- summarise(by_weekdays_and_interval, average.total.daily.steps = mean(steps, na.rm = TRUE))
```

2. Ploting the time series of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis):

```{r}
g <- ggplot(data = activityfill_by_intervals, aes(interval, average.total.daily.steps))
g + geom_line() + facet_wrap(~weekDay, nrow = 2) + xlab("5 minutes intervals") +
        ggtitle("Time Series of average total steps \n per day by interval")
```

**Conclusion**: the weekdays and weekend activity patterns  are slightly different. They share a maximum intensity around interval 515 (08h00 - 08h05) but in the afternoon and evening, the weekend activity is globally more intense.